Towards Practical First-Order Model Counting

### 2025-07-06 doc/talk/talk.tex
More slides

% TODO: 5 slides explaining the solution (maybe including the theorems)
% 0. Preprocess
% 1. Use Crane to compile into functions
% 2. Simplify
% 3. FindBaseCases TODO
% 4. For each base case
% 4.1. For each constant (TODO?)
% 4.1.1. Propagate TODO
% 5. Compile to C++

% TODO: (here and on the next slide) mark which steps will be explained in more detail
\begin{frame}{Knowledge Compilation Workflow (1/2)}
  \begin{enumerate}
    \item Use \textsc{Crane} to compile \structure{$\phi$} into a set of
          equations \structure{$\mathcal{E}$} \pause
    \item Simplify them, e.g.,
          \[
            g(l, m) = \sum_{k=0}^{m}[0 \le k \le 1]\binom{m}{k}g(l-1, m-k)
          \]
          becomes
          \[
            g(l, m) = g(l-1, m) + mg(l-1, m-1)
          \] \pause
    \item Identify a sufficient set of base cases
          \begin{itemize}
            \item e.g., \structure{$\{\, g(0, m), g(l, 0) \,\}$}
          \end{itemize}
  \end{enumerate}
\end{frame}

@@

  \begin{tikzcd}
    g(0, m) \arrow[d, gray, visible on=<2->] & \phantom{-} \arrow[d, phantom, visible on=<2->, "\text{\structure{4.1.} Construct the corr.\ sentence}"] & g(l, 0) \arrow[d, gray, visible on=<2->] \\
    |[visible on=<2->]| \forall x \in \Gamma\text{. }S(x) \lor \neg S(x) \arrow[d, gray, visible on=<3->] & \phantom{-} \arrow[d, phantom, visible on=<3->, "\text{\structure{4.2.} Recurse}"] & |[visible on=<2->]| \top \arrow[d, gray, visible on=<3->] \\
    |[visible on=<3->]| g(0, m) = \alt<4->{0^m}{\text{???}} \arrow[rd, gray, visible on=<5->] & |[visible on=<5->]| \text{\structure{4.3.} Add to }\structure{\mathcal{E}} & |[visible on=<3->]| g(l, 0) = \alt<4->{1}{???} \arrow[ld, gray, visible on=<5->] \\
    & |[visible on=<5->]| \mathcal{E} \arrow[d, gray, visible on=<6->, "\text{\color{black}{\structure{5.} Compile to C++}}"] & \\
    |[visible on=<7->]| \text{Domain sizes} \arrow[r, gray, visible on=<7->] & |[visible on=<6->]| \text{\dbox{C++ code}} \arrow[r, gray, visible on=<7->] & |[visible on=<7->]| \text{Model count}
  \end{tikzcd}
\end{frame}

>>>>

\begin{frame}{Knowledge Compilation Workflow (1/2)}
  \begin{enumerate}
    \item Use \textsc{Crane} to compile \structure{$\phi$} into a set of
          equations \structure{$\mathcal{E}$} \pause
    \item Simplify them, e.g.,
          \[
            g(l, m) = \sum_{k=0}^{m}[0 \le k \le 1]\binom{m}{k}g(l-1, m-k)
          \]
          becomes
          \[
            g(l, m) = g(l-1, m) + mg(l-1, m-1)
          \] \pause
    \item[3. ($\Rightarrow$)] Identify a sufficient set of base cases
          \begin{itemize}
            \item e.g., \structure{$\{\, g(0, m), g(l, 0) \,\}$}
          \end{itemize}
  \end{enumerate}
\end{frame}
% NOTE: explain what the arrow means

@@

  \begin{tikzcd}
    g(0, m) \arrow[d, gray, visible on=<2->] & \phantom{-} \arrow[d, phantom, visible on=<2->, "\text{\structure{4.1. ($\Rightarrow$)} Construct the corr.\ sentence}"] & g(l, 0) \arrow[d, gray, visible on=<2->] \\
    |[visible on=<2->]| \forall x \in \Gamma\text{. }S(x) \lor \neg S(x) \arrow[d, gray, visible on=<3->] & \phantom{-} \arrow[d, phantom, visible on=<3->, "\text{\structure{4.2.} Recurse}"] & |[visible on=<2->]| \top \arrow[d, gray, visible on=<3->] \\
    |[visible on=<3->]| g(0, m) = \alt<4->{0^m}{\text{???}} \arrow[rd, gray, visible on=<5->] & |[visible on=<5->]| \text{\structure{4.3.} Add to }\structure{\mathcal{E}} & |[visible on=<3->]| g(l, 0) = \alt<4->{1}{???} \arrow[ld, gray, visible on=<5->] \\
    & |[visible on=<5->]| \mathcal{E} \arrow[d, gray, visible on=<6->, "\text{\color{black}{\structure{5.} Compile to C++}}"] & \\
    |[visible on=<7->]| \text{Domain sizes} \arrow[r, gray, visible on=<7->] & |[visible on=<6->]| \text{\dbox{C++ code}} \arrow[r, gray, visible on=<7->] & |[visible on=<7->]| \text{Model count}
  \end{tikzcd}
\end{frame}

\begin{frame}[fragile]{Finding (a Sufficient Set of) Base Cases}
  \begin{block}{Outline}
    \begin{enumerate}
      \item For every function call:
            \begin{enumerate}
              \item For every argument of the form
                    \structure{$\textit{var} - \textit{const}$}:
                    \begin{enumerate}
                      \item Replace the signature parameter with
                            \structure{$0$}, \structure{$1$}, \dots,
                            \structure{$\textit{const} - 1$}
                    \end{enumerate}
              \item For every argument of the form \structure{$\textit{const}$}:
                    \begin{enumerate}
                      \item Replace the corresponding signature parameter with
                            \structure{$\textit{const}$}
                    \end{enumerate}
            \end{enumerate}
    \end{enumerate}
  \end{block}
  \begin{example}
    The signature of \structure{$g$} is \structure{$g(l, m)$}.

    \centering
    \begin{tikzcd}
      \text{Function calls:} & g(l-1, m) \arrow[d] & g(l-1, m-1) \arrow[ld] \arrow[d] \\
      \text{Base cases:} & g(0, m) & g(l, 0)
    \end{tikzcd}
  \end{example}
\end{frame}

\begin{frame}{Theoretical Results}
  \begin{theorem}
    The maximum \alert{recursion depth} of the compilation algorithm is upper
    bounded by the \alert{number of domains}.
  \end{theorem}
  \pause
  \begin{proof}[Proof (hint).]
    Each recursive call eliminates a domain.
  \end{proof}
  \pause
  \begin{theorem}
    The \alert{evaluation} of a recursive function always \alert{terminates}.
  \end{theorem}
  \pause
  \begin{proof}[Proof (hints).]
    \begin{itemize}
      \item There exists a \alert{topological ordering} of functions
      \item All function calls follow the \alert{structure} from the previous
            slide
      \item Some common-sense assumptions about the \alert{evaluation order} and
            previous work
    \end{itemize}
  \end{proof}
\end{frame}

% TODO: propagate + maybe smoothing
\begin{frame}{From a Base Case to a Sentence}
  \begin{block}{Preliminaries}
    \begin{itemize}
      \item \textsc{Crane} associates each function \structure{$f$} with a
            sentence \structure{$\phi$} such that
            \structure{$\textsc{Crane}(\phi)$} produces the definition of
            \structure{$f$}
      \item There is a bijection between the parameters of \structure{$f$} and
            the domains of \structure{$\phi$}
    \end{itemize}
  \end{block}
\end{frame}
